---
title: "Retirement Portfolio Monte Carlo Simulation"
author: "Thomas Benacci"
date: "2025-11-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Libraries

```{r}
library(tidyverse)
library(ggplot2)
```

# Load Personalized Mortality Profile

```{r}
mort <- readRDS("F:/PredictingMortality/mc_mortality_input.rds")

cat("Current age:            ", mort$current_age, "\n")
cat("Expected death age:     ", round(mort$expected_death_age, 1), "\n")
cat("Median death age:       ", round(mort$median_death_age, 1), "\n")
cat("Std dev:                ", round(mort$sd_death_age, 1), "\n")
cat("\nPercentiles:\n")
print(mort$percentiles)
```

# ============================================================
# YOUR FINANCIAL INPUTS
# All dollar amounts are in TODAY'S dollars.
# ============================================================

## Income & Career

```{r}
current_age       <- mort$current_age
retirement_age    <- 65       # When you plan to stop working
current_income    <- 75000    # Current annual gross income (today's $)
real_income_growth <- 0.02    # Real annual income growth (above inflation)
savings_rate      <- 0.20     # Fraction of gross income saved per year
```

## Portfolio

```{r}
current_portfolio <- 50000    # Current total retirement savings (today's $)
stock_allocation  <- 0.70     # Fraction in equities (rest in bonds)
# Rebalanced annually to this target
```

## Retirement Spending

```{r}
# Annual spending in retirement (today's dollars)
# This will grow with realized inflation each year in the simulation.
annual_spending   <- 50000

# Social Security or pension (today's dollars, COLA-adjusted with inflation)
social_security   <- 24000    # Annual SS benefit
ss_start_age      <- 67       # Age SS payments begin
```

## Return & Inflation Assumptions

Returns are NOMINAL. Inflation is modeled as a separate stochastic process,
correlated with stock and bond returns. Real values are computed per-path
by deflating nominal values by each path's realized cumulative inflation.

Historical calibration (US 1926-2023, approximate):

```{r}
# Nominal stock parameters
stock_mean_nom <- 0.10     # Expected nominal return
stock_sd       <- 0.19     # Annual volatility

# Nominal bond parameters (intermediate-term)
bond_mean_nom  <- 0.05     # Expected nominal return
bond_sd        <- 0.07     # Annual volatility

# Inflation parameters
infl_mean      <- 0.03     # Expected annual inflation
infl_sd        <- 0.02     # Inflation volatility (historically ~1.5-2.5%)

# Correlations (approximate historical)
corr_stock_bond <- 0.10    # Stocks and bonds
corr_stock_infl <- 0.02    # Stocks and inflation (weak positive)
corr_bond_infl  <- -0.30   # Bonds and inflation (negative: bonds hurt by inflation)
```

## Simulation Parameters

```{r}
n_sims       <- 10000   # Number of Monte Carlo simulations
max_age      <- 105     # Hard ceiling
set.seed(42)
```

# ============================================================
# SIMULATION ENGINE
# ============================================================

## Sample Death Ages

```{r}
sample_death_age <- function(n, mort_table) {
  u <- runif(n)
  sapply(u, function(p) {
    idx <- which(mort_table$cum_death_prob >= p)[1]
    if (is.na(idx)) return(max(mort_table$age))
    mort_table$age[idx]
  })
}

death_ages <- sample_death_age(n_sims, mort$mortality_table)

cat("Sampled death ages:\n")
cat("  Mean:   ", round(mean(death_ages), 1), "\n")
cat("  Median: ", round(median(death_ages), 1), "\n")
cat("  Min:    ", min(death_ages), "  Max:", max(death_ages), "\n")
```

## Build 3x3 Covariance Matrix and Cholesky Factor

```{r}
# Order: stocks, bonds, inflation
corr_matrix <- matrix(c(
  1,                corr_stock_bond,  corr_stock_infl,
  corr_stock_bond,  1,                corr_bond_infl,
  corr_stock_infl,  corr_bond_infl,   1
), nrow = 3, byrow = TRUE)

sd_vec <- c(stock_sd, bond_sd, infl_sd)
cov_matrix <- diag(sd_vec) %*% corr_matrix %*% diag(sd_vec)

# Verify positive-definiteness
stopifnot(all(eigen(cov_matrix)$values > 0))

chol_L <- chol(cov_matrix)  # Upper Cholesky: z %*% chol_L gives correlated draws

# Log-normal drift adjustments (so E[exp(X)] matches desired arithmetic mean)
stock_drift <- log(1 + stock_mean_nom) - stock_sd^2 / 2
bond_drift  <- log(1 + bond_mean_nom)  - bond_sd^2 / 2
infl_drift  <- log(1 + infl_mean)      - infl_sd^2 / 2

generate_paths <- function(n_years) {
  z <- matrix(rnorm(n_years * 3), ncol = 3)
  corr_z <- z %*% chol_L
  data.frame(
    stock_r = exp(stock_drift + corr_z[, 1]) - 1,
    bond_r  = exp(bond_drift  + corr_z[, 2]) - 1,
    infl_r  = exp(infl_drift  + corr_z[, 3]) - 1
  )
}

cat("Covariance matrix (stocks, bonds, inflation):\n")
print(round(cov_matrix, 5))
```

## Run Simulation

```{r}
n_years_max <- max_age - current_age

# Storage: nominal portfolio, real portfolio, and cumulative price level
nominal_matrix  <- matrix(NA, nrow = n_sims, ncol = n_years_max + 1)
real_matrix     <- matrix(NA, nrow = n_sims, ncol = n_years_max + 1)
price_level_mat <- matrix(NA, nrow = n_sims, ncol = n_years_max + 1)
ruin_age        <- rep(NA, n_sims)

for (i in 1:n_sims) {
  
  portfolio_nom <- current_portfolio
  income_nom    <- current_income
  spending_nom  <- annual_spending
  ss_nom        <- social_security
  price_level   <- 1.0          # Cumulative price level (1 = today's dollars)
  death         <- death_ages[i]
  
  paths <- generate_paths(n_years_max)
  
  for (t in 1:n_years_max) {
    age <- current_age + t - 1
    if (age >= death) break
    
    # Record state at start of year
    nominal_matrix[i, t]  <- portfolio_nom
    price_level_mat[i, t] <- price_level
    real_matrix[i, t]     <- portfolio_nom / price_level
    
    # This year's realized inflation
    infl_this_year <- paths$infl_r[t]
    
    # --- Cash flows (all nominal) ---
    if (age < retirement_age) {
      contribution_nom <- income_nom * savings_rate
      withdrawal_nom   <- 0
      income_nom <- income_nom * (1 + infl_this_year) * (1 + real_income_growth)
    } else {
      ss_this_year <- ifelse(age >= ss_start_age, ss_nom, 0)
      withdrawal_nom   <- max(spending_nom - ss_this_year, 0)
      contribution_nom <- 0
    }
    
    # Nominal portfolio return (blended)
    r_blend <- stock_allocation * paths$stock_r[t] +
               (1 - stock_allocation) * paths$bond_r[t]
    
    # Update nominal portfolio
    portfolio_nom <- portfolio_nom - withdrawal_nom
    
    if (portfolio_nom <= 0) {
      portfolio_nom <- 0
      if (is.na(ruin_age[i])) ruin_age[i] <- age
      nominal_matrix[i, t] <- 0
      real_matrix[i, t]    <- 0
      price_level  <- price_level * (1 + infl_this_year)
      spending_nom <- spending_nom * (1 + infl_this_year)
      ss_nom       <- ss_nom * (1 + infl_this_year)
      next
    }
    
    portfolio_nom <- portfolio_nom * (1 + r_blend) + contribution_nom
    portfolio_nom <- max(portfolio_nom, 0)
    
    # Advance price level and nominal cash flows
    price_level  <- price_level * (1 + infl_this_year)
    spending_nom <- spending_nom * (1 + infl_this_year)
    ss_nom       <- ss_nom * (1 + infl_this_year)
  }
  
  # Record final values
  final_t <- min(death - current_age, n_years_max)
  if (final_t > 0 && final_t <= ncol(nominal_matrix)) {
    nominal_matrix[i, final_t]  <- portfolio_nom
    price_level_mat[i, final_t] <- price_level
    real_matrix[i, final_t]     <- portfolio_nom / price_level
  }
}

cat("Simulation complete:", n_sims, "runs\n")
```

# ============================================================
# RESULTS (all reported in today's dollars unless noted)
# ============================================================

# Probability of Ruin

```{r}
n_ruined <- sum(!is.na(ruin_age))
prob_ruin <- n_ruined / n_sims

cat("=== PROBABILITY OF RUIN ===\n")
cat("Portfolio depleted before death: ", n_ruined, "/", n_sims, "\n")
cat("Probability of ruin:             ", sprintf("%.1f%%", prob_ruin * 100), "\n")
cat("Probability of success:          ", sprintf("%.1f%%", (1 - prob_ruin) * 100), "\n\n")

if (n_ruined > 0) {
  cat("Among ruined scenarios:\n")
  cat("  Mean ruin age:    ", round(mean(ruin_age, na.rm = TRUE), 1), "\n")
  cat("  Median ruin age:  ", round(median(ruin_age, na.rm = TRUE), 1), "\n")
  cat("  Earliest ruin:    ", min(ruin_age, na.rm = TRUE), "\n")
  cat("  Years in ruin:    ", round(mean(death_ages[!is.na(ruin_age)] - ruin_age[!is.na(ruin_age)]), 1), "avg\n")
}
```

# Inflation Paths

```{r, fig.width=9, fig.height=5}
pct_infl <- do.call(rbind, lapply(1:n_years_max, function(t) {
  vals <- price_level_mat[, t]
  vals <- vals[!is.na(vals)]
  if (length(vals) < 10) return(NULL)
  data.frame(
    age = current_age + t - 1,
    p10 = quantile(vals, 0.10), p25 = quantile(vals, 0.25),
    p50 = quantile(vals, 0.50), p75 = quantile(vals, 0.75),
    p90 = quantile(vals, 0.90)
  )
}))

ggplot(pct_infl, aes(x = age)) +
  geom_ribbon(aes(ymin = p10, ymax = p90), fill = "darkorange", alpha = 0.15) +
  geom_ribbon(aes(ymin = p25, ymax = p75), fill = "darkorange", alpha = 0.25) +
  geom_line(aes(y = p50), color = "darkorange", linewidth = 1.2) +
  labs(title = "Cumulative Price Level (Stochastic Inflation)",
       subtitle = "1.0 = today's purchasing power | Median, IQR, and 10-90 bands",
       x = "Age", y = "Price Level (1 = today)") +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold"))
```

# Portfolio Value Trajectories (Today's Dollars)

```{r, fig.width=10, fig.height=7}
pct_real <- do.call(rbind, lapply(1:n_years_max, function(t) {
  vals <- real_matrix[, t]
  vals <- vals[!is.na(vals)]
  if (length(vals) < 10) return(NULL)
  data.frame(
    age = current_age + t - 1,
    p10 = quantile(vals, 0.10), p25 = quantile(vals, 0.25),
    p50 = quantile(vals, 0.50), p75 = quantile(vals, 0.75),
    p90 = quantile(vals, 0.90)
  )
}))

pct_nom <- do.call(rbind, lapply(1:n_years_max, function(t) {
  vals <- nominal_matrix[, t]
  vals <- vals[!is.na(vals)]
  if (length(vals) < 10) return(NULL)
  data.frame(age = current_age + t - 1, p50_nom = quantile(vals, 0.50))
}))

ggplot(pct_real, aes(x = age)) +
  geom_ribbon(aes(ymin = p10, ymax = p90), fill = "steelblue", alpha = 0.15) +
  geom_ribbon(aes(ymin = p25, ymax = p75), fill = "steelblue", alpha = 0.25) +
  geom_line(aes(y = p50), color = "steelblue", linewidth = 1.2) +
  geom_line(data = pct_nom, aes(x = age, y = p50_nom),
            color = "grey50", linewidth = 0.8, linetype = "dashed") +
  geom_vline(xintercept = retirement_age, linetype = "dashed", color = "grey40") +
  annotate("text", x = retirement_age + 0.5, y = max(pct_real$p90, na.rm = TRUE) * 0.95,
           label = "Retirement", hjust = 0, size = 3.5) +
  geom_hline(yintercept = 0, color = "firebrick", linewidth = 0.5) +
  scale_y_continuous(labels = scales::dollar_format()) +
  labs(title = "Portfolio Value Over Time (Today's Dollars)",
       subtitle = paste0("Blue = real (inflation-adjusted) | Dashed grey = nominal median | ",
                         sprintf("P(ruin) = %.1f%%", prob_ruin * 100)),
       x = "Age", y = "Portfolio Value (today's $)") +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold"))
```

# Terminal Wealth Distribution (Today's Dollars)

```{r, fig.width=9, fig.height=5}
terminal_real <- sapply(1:n_sims, function(i) {
  t_death <- min(death_ages[i] - current_age, n_years_max)
  if (t_death < 1) return(0)
  val <- real_matrix[i, t_death]
  ifelse(is.na(val), 0, val)
})

terminal_nom <- sapply(1:n_sims, function(i) {
  t_death <- min(death_ages[i] - current_age, n_years_max)
  if (t_death < 1) return(0)
  val <- nominal_matrix[i, t_death]
  ifelse(is.na(val), 0, val)
})

cat("=== TERMINAL WEALTH AT DEATH ===\n\n")
cat("In today's dollars (real):\n")
cat("  Mean:    ", scales::dollar(mean(terminal_real)), "\n")
cat("  Median:  ", scales::dollar(median(terminal_real)), "\n")
cat("  10th %:  ", scales::dollar(quantile(terminal_real, 0.10)), "\n")
cat("  90th %:  ", scales::dollar(quantile(terminal_real, 0.90)), "\n\n")
cat("In nominal dollars:\n")
cat("  Mean:    ", scales::dollar(mean(terminal_nom)), "\n")
cat("  Median:  ", scales::dollar(median(terminal_nom)), "\n")
cat("  10th %:  ", scales::dollar(quantile(terminal_nom, 0.10)), "\n")
cat("  90th %:  ", scales::dollar(quantile(terminal_nom, 0.90)), "\n")

cap <- quantile(terminal_real[terminal_real > 0], 0.99, na.rm = TRUE)
if (is.na(cap) || cap <= 0) cap <- 1
ggplot(data.frame(w = pmin(terminal_real, cap)), aes(x = w)) +
  geom_histogram(bins = 60, fill = "steelblue", alpha = 0.7) +
  geom_vline(xintercept = 0, color = "firebrick", linewidth = 0.8) +
  scale_x_continuous(labels = scales::dollar_format()) +
  labs(title = "Distribution of Wealth at Death (Today's Dollars)",
       subtitle = paste0("Across ", scales::comma(n_sims),
                         " simulations | personalized mortality & stochastic inflation"),
       x = "Portfolio Value at Death (today's $)", y = "Count") +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold"))
```

# Withdrawal Rate Sensitivity

Spending is specified in today's dollars and grows with each path's realized inflation.

```{r, fig.width=9, fig.height=6}
spending_levels <- seq(30000, 80000, by = 5000)
ruin_rates <- numeric(length(spending_levels))

# Pre-generate paths for consistency across sensitivity tests
all_paths <- lapply(1:n_sims, function(i) generate_paths(n_years_max))

for (s in seq_along(spending_levels)) {
  test_spending_real <- spending_levels[s]
  test_ruin <- 0
  
  for (i in 1:n_sims) {
    portfolio_nom <- current_portfolio
    income_nom    <- current_income
    spending_nom  <- test_spending_real
    ss_nom        <- social_security
    death         <- death_ages[i]
    paths         <- all_paths[[i]]
    ruined        <- FALSE
    
    for (t in 1:n_years_max) {
      age <- current_age + t - 1
      if (age >= death) break
      infl <- paths$infl_r[t]
      
      if (age < retirement_age) {
        contribution_nom <- income_nom * savings_rate
        withdrawal_nom   <- 0
        income_nom <- income_nom * (1 + infl) * (1 + real_income_growth)
      } else {
        ss_this_year   <- ifelse(age >= ss_start_age, ss_nom, 0)
        withdrawal_nom <- max(spending_nom - ss_this_year, 0)
        contribution_nom <- 0
      }
      
      portfolio_nom <- portfolio_nom - withdrawal_nom
      if (portfolio_nom <= 0) { ruined <- TRUE; break }
      
      r_blend <- stock_allocation * paths$stock_r[t] +
                 (1 - stock_allocation) * paths$bond_r[t]
      portfolio_nom <- portfolio_nom * (1 + r_blend) + contribution_nom
      
      spending_nom <- spending_nom * (1 + infl)
      ss_nom       <- ss_nom * (1 + infl)
    }
    if (ruined) test_ruin <- test_ruin + 1
  }
  ruin_rates[s] <- test_ruin / n_sims
}

ruin_frontier <- data.frame(
  annual_spending = spending_levels,
  withdrawal_rate_pct = spending_levels / current_portfolio * 100,
  prob_ruin = ruin_rates
)

ggplot(ruin_frontier, aes(x = annual_spending, y = prob_ruin)) +
  geom_line(color = "firebrick", linewidth = 1.2) +
  geom_point(color = "firebrick", size = 2) +
  geom_hline(yintercept = 0.05, linetype = "dashed", color = "grey40") +
  annotate("text", x = min(spending_levels), y = 0.07,
           label = "5% ruin threshold", hjust = 0, size = 3.5, color = "grey40") +
  geom_hline(yintercept = 0.10, linetype = "dotted", color = "grey60") +
  annotate("text", x = min(spending_levels), y = 0.12,
           label = "10% ruin threshold", hjust = 0, size = 3.5, color = "grey60") +
  scale_x_continuous(labels = scales::dollar_format()) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, NA)) +
  labs(title = "Probability of Ruin by Annual Spending Level (Today's $)",
       subtitle = paste0("Retirement at ", retirement_age,
                         " | SS $", scales::comma(social_security), "/yr at age ", ss_start_age,
                         " | ", round(stock_allocation*100), "/", round((1-stock_allocation)*100),
                         " stocks/bonds | stochastic inflation"),
       x = "Annual Spending (today's $)", y = "Probability of Ruin") +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold"))

safe_5  <- ruin_frontier %>% filter(prob_ruin <= 0.05) %>% slice_max(annual_spending, n = 1)
safe_10 <- ruin_frontier %>% filter(prob_ruin <= 0.10) %>% slice_max(annual_spending, n = 1)
cat("\nMax annual spending at <= 5% ruin: ",
    ifelse(nrow(safe_5) > 0, scales::dollar(safe_5$annual_spending), "none in range"), "\n")
cat("Max annual spending at <= 10% ruin:",
    ifelse(nrow(safe_10) > 0, scales::dollar(safe_10$annual_spending), "none in range"), "\n")
```

# Retirement Age Sensitivity

```{r, fig.width=9, fig.height=6}
ret_ages <- seq(55, 75, by = 5)
ruin_by_ret_age <- numeric(length(ret_ages))

for (r in seq_along(ret_ages)) {
  test_ret_age <- ret_ages[r]
  test_ruin <- 0
  for (i in 1:n_sims) {
    portfolio_nom <- current_portfolio
    income_nom    <- current_income
    spending_nom  <- annual_spending
    ss_nom        <- social_security
    death         <- death_ages[i]
    paths         <- all_paths[[i]]
    ruined        <- FALSE
    for (t in 1:n_years_max) {
      age <- current_age + t - 1
      if (age >= death) break
      infl <- paths$infl_r[t]
      if (age < test_ret_age) {
        contribution_nom <- income_nom * savings_rate
        withdrawal_nom   <- 0
        income_nom <- income_nom * (1 + infl) * (1 + real_income_growth)
      } else {
        ss_this_year   <- ifelse(age >= ss_start_age, ss_nom, 0)
        withdrawal_nom <- max(spending_nom - ss_this_year, 0)
        contribution_nom <- 0
      }
      portfolio_nom <- portfolio_nom - withdrawal_nom
      if (portfolio_nom <= 0) { ruined <- TRUE; break }
      r_blend <- stock_allocation * paths$stock_r[t] +
                 (1 - stock_allocation) * paths$bond_r[t]
      portfolio_nom <- portfolio_nom * (1 + r_blend) + contribution_nom
      spending_nom <- spending_nom * (1 + infl)
      ss_nom       <- ss_nom * (1 + infl)
    }
    if (ruined) test_ruin <- test_ruin + 1
  }
  ruin_by_ret_age[r] <- test_ruin / n_sims
}

ggplot(data.frame(ret_age = ret_ages, prob_ruin = ruin_by_ret_age),
       aes(x = ret_age, y = prob_ruin)) +
  geom_line(color = "darkgreen", linewidth = 1.2) +
  geom_point(color = "darkgreen", size = 3) +
  scale_y_continuous(labels = scales::percent_format(), limits = c(0, NA)) +
  labs(title = "Probability of Ruin by Retirement Age",
       subtitle = paste0("Spending $", scales::comma(annual_spending),
                         "/yr (today's $) | stochastic inflation"),
       x = "Retirement Age", y = "Probability of Ruin") +
  theme_minimal() +
  theme(plot.title = element_text(size = 16, face = "bold"))
```

# Summary

```{r}
median_cum_infl <- median(sapply(1:n_sims, function(i) {
  t_death <- min(death_ages[i] - current_age, n_years_max)
  if (t_death < 1) return(1)
  pl <- price_level_mat[i, t_death]
  ifelse(is.na(pl), 1, pl)
}))

cat("============================================\n")
cat("  RETIREMENT MONTE CARLO SIMULATION SUMMARY\n")
cat("============================================\n\n")
cat("Inputs:\n")
cat("  Current age:          ", current_age, "\n")
cat("  Retirement age:       ", retirement_age, "\n")
cat("  Current portfolio:    ", scales::dollar(current_portfolio), "(today's $)\n")
cat("  Current income:       ", scales::dollar(current_income), "(today's $)\n")
cat("  Savings rate:         ", sprintf("%.0f%%", savings_rate * 100), "\n")
cat("  Annual spending:      ", scales::dollar(annual_spending), "(today's $, grows w/ inflation)\n")
cat("  Social Security:      ", scales::dollar(social_security), "/yr at age", ss_start_age, "(COLA)\n")
cat("  Allocation:           ", sprintf("%.0f/%.0f stocks/bonds",
    stock_allocation*100, (1-stock_allocation)*100), "\n\n")
cat("Market Assumptions (NOMINAL):\n")
cat("  Stock return:         ", sprintf("%.1f%% +/- %.1f%%", stock_mean_nom*100, stock_sd*100), "\n")
cat("  Bond return:          ", sprintf("%.1f%% +/- %.1f%%", bond_mean_nom*100, bond_sd*100), "\n")
cat("  Inflation:            ", sprintf("%.1f%% +/- %.1f%%", infl_mean*100, infl_sd*100), "\n")
cat("  Stock-bond corr:      ", corr_stock_bond, "\n")
cat("  Stock-inflation corr: ", corr_stock_infl, "\n")
cat("  Bond-inflation corr:  ", corr_bond_infl, "\n\n")
cat("Mortality (personalized Cox PH model):\n")
cat("  Expected death age:   ", round(mort$expected_death_age, 1), "\n")
cat("  Median death age:     ", round(mort$median_death_age, 1), "\n")
cat("  Std dev:              ", round(mort$sd_death_age, 1), "years\n\n")
cat("Results (", scales::comma(n_sims), " simulations):\n")
cat("  Probability of ruin:      ", sprintf("%.1f%%", prob_ruin * 100), "\n")
cat("  Median terminal (real):   ", scales::dollar(median(terminal_real)), "\n")
cat("  Mean terminal (real):     ", scales::dollar(mean(terminal_real)), "\n")
cat("  Median terminal (nominal):", scales::dollar(median(terminal_nom)), "\n")
cat("  Median cumulative infl:   ", sprintf("%.1fx (%.0f%% total)",
    median_cum_infl, (median_cum_infl-1)*100), "\n")
if (exists("safe_5") && nrow(safe_5) > 0)
  cat("  Max safe spend (5%):  ", scales::dollar(safe_5$annual_spending), "(today's $)\n")
if (exists("safe_10") && nrow(safe_10) > 0)
  cat("  Max safe spend (10%): ", scales::dollar(safe_10$annual_spending), "(today's $)\n")
```

---

**Methodology notes:**

- **Stochastic inflation**: Inflation is a correlated lognormal process
  (mean `r sprintf("%.1f%%", infl_mean*100)`, SD `r sprintf("%.1f%%", infl_sd*100)`),
  jointly drawn with stock and bond returns via a 3x3 Cholesky decomposition.
  Each simulation path realizes its own inflation trajectory.
- **Nominal simulation, real reporting**: Portfolio, income, spending, and Social
  Security evolve in nominal terms. All reported statistics are deflated by each
  path's own cumulative price level, so "today's dollars" means something different
  for each path — correctly capturing the fact that high-inflation paths erode
  purchasing power more.
- **Key correlation**: Bond-inflation correlation is negative (-0.30), capturing
  the empirical fact that surprise inflation hurts nominal bond holders. This means
  the 70/30 portfolio faces correlated risk: high inflation simultaneously erodes
  bond values and increases the nominal spending needed to maintain real lifestyle.
- **Mortality** is sampled from the personalized Cox PH survival curve, not generic
  actuarial tables. Each simulation jointly draws a death age, a market path, and
  an inflation path.
- Following Stout & Mitchell (2006, 2014), Milevsky & Robinson (2005), and
  addressing the inflation-risk concerns of Pfau (2010) and Bengen (2012).
